### 2020-10-15 priority-gv.tex
on typing in pgv

\subsubsection*{Typing}
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$, which states that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing is standard for the linear $\lambda$-calculus. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. The constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const}. \Cref{fig:pgv-typing} gives the typing rules for PGV.

@@

In PGV, sequents are annotated by a priority bound $\cs{p}$, which is the \emph{upper bound} on the priority of the communication actions performed while evaluating the term. The \emph{lower bound} is obtained from the typing environment. Terms with sequential evaluation, such as \LabTirName{T-LetUnit}, have side conditions to ensure deadlock freedom. For instance, the side condition for typing $\tm{\letunit{M}{N}}$ requires that $\tm{M}$ has finished communicating before $\tm{N}$ starts communicating. The rule \LabTirName{T-Lam} stores the lower and upper priority bound of the function body in the type.

Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$, which states that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well-typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensure that there is at most one main thread.

While our configuration typing is based on the standard typing for GV, it differs on two counts.
Firstly, we require that child threads return the unit value, as opposed to a terminated channel.
Secondly, we simplify typing for parallel composition. In GV, each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$ across the parallel composition. Consequently, associativity of parallel composition does not preserve typing. Unfortunately, splitting exactly one channel across a parallel composition in GV plays a crucial role in guaranteeing deadlock freedom. In PGV, we guarantee deadlock freedom using priorities, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.

>>>>

\subsubsection*{Typing}
\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}. Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}).

@@

In PGV, sequents are annotated with a priority bound $\cs{p}$, which is the \emph{upper bound} on the priority of actions performed while evaluating the term. The \emph{lower bound} is obtained from the typing environment. Terms with sequential evaluation, such as \LabTirName{T-LetUnit}, have side conditions to ensure deadlock freedom. For instance, the side condition for typing $\tm{\letunit{M}{N}}$ requires that $\tm{M}$ has finished communicating before $\tm{N}$ starts communicating. The rule \LabTirName{T-Lam} stores the lower and upper priority bound of the function body in the type.

Typing rules for configurations are at the bottom of \cref{fig:pgv-typing}. Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ stating that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensure that there is at most one main thread.

While our configuration typing is based on the standard typing for GV, it differs on two counts
$(i)$ we require that child threads return the unit value, as opposed to a terminated channel;
$(ii)$ we simplify typing for parallel composition.
In order to guarantee deadlock freedom, in GV each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$. Consequently, associativity of parallel composition does not preserve typing. In PGV, we guarantee deadlock freedom using priorities, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.
